### 2024-04-03 paper/paper-papaja.Rmd
paper updates

## Public Significance Statement

It is well-established that mere exposure to statements, such as in a true-or-false quiz, can lead people to later rate those same statements as more true (the “illusory truth” effect). On most accounts, the illusory truth effect is thought to be caused by fluency or familiarity with a specific statement, and therefore should not more widely affect people's beliefs or decisions in most cases. In contrast to these prior accounts, however, we found that mere exposure to statements affected participants' responses to other novel statements (what we term an “illusory implication” effect), suggesting that mere exposure can instead lead to genuine belief. Our findings are further cause for concern about the spread and influence of misinformation, as they indicate that any exposure to misinformation may cause harm by inducing false beliefs.

@@

@@

@@

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 election” might imply the belief that “Donald Trump actually won the 2020 election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

Generally, it is these wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then these effects may not have much impact outside the lab.  Under these theoretical accounts, the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar statement. This would make the real-world consequences of the illusory truth effect far narrower than one might fear: These metacognitive mechanisms are liable to apply only in response to the exact statements previously seen, but are not as likely to generalize to other related ideas, or to drive actions or decision-making in novel contexts.

@@

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related statements.

A handful of studies are suggestive in this regard: Unkelbach and Rom [-@unkelbach.rom2017] found that mere exposure to one statement could increase endorsements of implied statements. For instance, being more likely to endorse the statement "Most accidents occur close to weekends" after exposure to the statement "Most accidents occur on Mondays". Similarly, Arkes and colleagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, in both cases these findings can also be easily explained by fluency or familiarity engendered by the similarity between the presented and tested statements [@arkes.etal1991; @unkelbach.rom2017].

Bacon [-@bacon1979] found that mere exposure to one statement ($P$) could lead to lower truth ratings for a new, contradictory statements($\neg P$). For instance, mere exposure to "crocodiles sleep with their eyes open" produced lower truth ratings for "crocodiles sleep with their eyes closed". It seems clear at least that familiarity cannot explain this pattern of results. However, other findings under very similar paradigms have partly contradicted these results [@garcia-marques.etal2015] and it is somewhat unclear whether fluency could account for these findings, i.e. whether contradictory statements can be expected to lead to fluent or disfluent processing.

Here, we sought to decisively examine whether mere exposure to statements could influence people's beliefs. To do so, we examined whether people draw inferences from statements after exposure in a context without any presupposition of truth. We designed a study to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for a  different statement the premise would logically entail (the “implication”). Such an “illusory implication” effect would indicate that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

In Experiments 1 and 2, the effect of familiarity was weaker than the measured implication effect, so that average truth ratings were higher after exposure for false-implied-true items and lower for true-implied-false items (Figures 1 and 2). In Experiment 3 however, the estimated effect of familiarity and implication were similar, and truth ratings were similar for exposed and unexposed true-implied-false items (Figure 3). This may suggest a possible alternate explanation of our findings: that the true-implied-false items' premise statements simply do not engender as much familiarity or fluency for their implication as do the false-implied-true items, so that the effects differ between these groups. Put another way, the identification of the "implication" effect coefficient rests on the assumption that familiarity and fluency effects are not accidentally confounded with item groups.

@@

What cognitive mechanism could explain these findings? One potential explanation is that mere exposure to the premise statements leads participants to at least partially adopt them as beliefs. Simply entertaining the idea may create such an impression despite awareness of its ambiguous truth-value [@gilbert.etal1990; @gilbert1991]. If so, this apparently occurs despite participant's awareness that the statements may be false, as ensured by constant reminders and comprehension checks in Experiment 3.

>>>>

@@

## Theories of the Illusory Truth Effect

@@

## Mere Exposure and Belief

@@

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

Generally, it is these wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then these effects may not have much impact outside the lab.  Under these theoretical accounts, the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar statement. This would make the real-world consequences of the illusory truth effect far narrower than one might otherwise fear: These metacognitive mechanisms are liable to apply only in response to the exact statements previously seen, but are not as likely to generalize to other related ideas, or to drive actions or decision-making in novel contexts.

@@

## Is there an "Illusory Implication" Effect?

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related statements. We refer to this hypothetical impact of mere exposure on implied statements as the "Illusory Implication" Effect.

A handful of studies are suggestive in this regard: Unkelbach and Rom [-@unkelbach.rom2017] found that mere exposure to one statement could increase endorsements of implied statements. For instance, people were more likely to endorse the statement "Most accidents occur close to weekends" after exposure to the statement "Most accidents occur on Mondays". Similarly, Arkes and colleagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, in both cases these findings can also be easily explained by fluency or familiarity engendered by the similarity between the presented and tested statements [@arkes.etal1991; @unkelbach.rom2017].

A case where mere exposure lead to a lower truth rating for a related statement (i.e., one implied to be false) would more clearly implicate an effect of mere exposure on beliefs. In our review of the literature, we find three reported experiments that offer preliminary evidence of this effect: Bacon [-@bacon1979] found that mere exposure to one statement ($P$) could lead to lower truth ratings for a new, contradictory statements($\neg P$) after a three-week delay. For instance, mere exposure to "crocodiles sleep with their eyes open" produced lower truth ratings for "crocodiles sleep with their eyes closed". Findings from an experiment by Garcia-Marques and colleagues [-@garcia-marques.etal2015] partially contradicted these results (using a slightly different exposure task), finding lower truth ratings for contradictory statements when participants were queried immediately, but finding the opposite pattern of results at the delay used by the original Bacon [-@bacon1979] study. A later series of experiments by Unkelbach and Rom [-@unkelbach.rom2017] replicated these new findings [@garcia-marques.etal2015], although the effects were not observed consistently in each experiment.

## The present studies

Here, we sought to expand upon prior experiments to decisively examine whether mere exposure to statements could influence people's beliefs by testing for an "illusory implication" effect. To do so, we examined whether people draw inferences from statements after exposure in a context without any presupposition of truth. We designed a series of studies to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for a  different statement the premise would logically entail (the “implication”). Crucially, we test this across cases where premises entail either the truth or falsehood of another related statement, ruling out purely metacognitive explanations.

In addition, we make several additional contributions toward definitively demonstrating the illusory implication effect beyond existing studies by 1) developing a new set of materials, 2) recruiting larger samples (present studies $N = 1020$, total prior cited studies $N = 452$), 3) utilizing modern hierarchical statistical modeling methods, and 4) taking great pains to ensure that participants' truth ratings are not influenced by task demands (Experiments 2 and 3). Across our studies, we find new compelling evidence for this “illusory implication” effect, indicating that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

In Experiments 1 and 2, the effect of familiarity was weaker than the measured implication effect, so that average truth ratings were higher after exposure for false-implied-true items and lower for true-implied-false items (Figures 1 and 2). In Experiment 3 however, the estimated effect of familiarity and implication were similar, and truth ratings were similar for exposed and unexposed true-implied-false items (Figure 3). Although the estimate for the implication effect in our statistical model was clearly greater than zero, this may nevertheless allow a possible alternate explanation of our findings: that the true-implied-false items' premise statements simply do not engender as much familiarity or fluency for their implication as do the false-implied-true items, so that the effects differ between these groups. Put another way, the identification of the "implication" effect coefficient rests on the assumption that familiarity and fluency effects are not accidentally confounded with item groups.

@@

What cognitive mechanism could explain these findings? One potential explanation is that mere exposure to the premise statements leads participants to at least partially adopt them as beliefs. Simply entertaining the idea may create such an impression despite awareness of its ambiguous truth-value [@gilbert.etal1990; @gilbert1991; @garcia-marques.etal2015]. If so, this apparently occurs despite participant's awareness that the statements may be false, as ensured by constant reminders and comprehension checks in Experiment 3.
